---
title: "Reproducible Research: Week 4 - Course Project 2"
author: "Joseph Boateng"
date: "1/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment: Reproducible Research: Week 4 - Course Project 2

## Overview

Weather events cause public health and economic problems for communities and municipalities. Severe events result in fatalities, injuries, and damage. Predicting and/or preventing these outcomes is a primary objective.

This analysis examines the damaging effects of severe weather conditions (e.g. hurricanes, tornadoes, thunderstorms, floods, etc.) on human populations and the econonomy in the U.S. from 1950 to 2011.

As a result, the analysis will highlight the severe weather events associated with the greatest impact on the economy and population health.

## Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

In this report,effect of weather events on personal as well as property damages was studied. Barplots were plotted seperately for the top 8 weather events that causes highest fatalities and highest injuries. Results indicate that most Fatalities and injuries were caused by Tornados.Also, barplots were plotted for the top 8 weather events that causes the highest property damage and crop damage.

 DATA PROCESSING 
 
 1. Data
The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

Storm Data [47Mb] There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

National Weather Service Storm Data Documentation

National Climatic Data Center Storm Events FAQ

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

## Assignment
The basic goal of this assignment is to explore the NOAA Storm Database and answer the following basic questions about severe weather events.
 
 Modus Operandi / Process
 
Data Preparation

1.1 Install packages & Load libraries

Install packages …

```{r, load libraries, echo=TRUE, cache=TRUE}
library(rmarkdown)
library(knitr)
#install.packages("R.utils",repos="http://cran.us.r-project.org")
library(R.utils)
 #install.packages("gridExtra", repos="'https://cran.rstudio.com")
 library(gridExtra)
#install.packages("ggplot2", repos="'https://cran.rstudio.com")
library(ggplot2)
```

1.2 Download the storm data file into the designated working directory folder

```{r,  Performing the download,echo=TRUE, cache=TRUE}
temp <- tempfile()
 
if(!file.exists("/stormData.csv.bz2")){
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile="./stormData.csv.bz2")
}
## Uncompressing the file
if(!file.exists("stormdata.csv"))
{
  bunzip2("stormData.csv.bz2","stormdata.csv",remove=F)
}

# 1.3 load & read data from file

storm <- read.csv("stormdata.csv",header=TRUE,sep=",")
summary(storm)
```

```{r, echo=TRUE}
names(storm)
```

RESULTS

QUESTION 1. 

Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

1.1 Variable selection (reducing the data set to only needed columns and variables)

```{r, echo=TRUE}

variables<-c("EVTYPE","FATALITIES","INJURIES","PROPDMG", "PROPDMGEXP","CROPDMG","CROPDMGEXP")

strmdata<-storm[variables]

dim(strmdata)

names(strmdata)
```

1.2 Reviewing events that cause the most fatalities ( The Top-10 Fatalities by Weather Event )

Procedure = aggregate the top 10 fatalities by the event type and sort the output in descending order

```{r, echo=TRUE}
Fatalities <- aggregate(FATALITIES ~ EVTYPE, data = strmdata, FUN = sum)

Top10_Fatalities <- Fatalities[order(-Fatalities$FATALITIES), ][1:10, ] 

Top10_Fatalities 

```

1.3 Reviewing events that cause the most injuries ( The Top-10 Injuries by Weather Event )

Procedure = aggregate the top 10 injuries by the event type and sort the output in descending order

```{r, echo=TRUE}

Injuries <- aggregate(INJURIES ~ EVTYPE, data = strmdata, FUN = sum)

Top10_Injuries <- Injuries[order(-Injuries$INJURIES), ][1:10, ] 

Top10_Injuries 

```

1.4 Plot of Top 10 Fatalities & Injuries for Weather Event Types ( Population Health Impact )

Proecedure = plot graphs showing the top 10 fatalities and injuries

```{r, ploting using gplot, echo=TRUE}

p1 = ggplot(Top10_Fatalities, aes(x = EVTYPE, y = FATALITIES, theme_set(theme_bw()))) + 
    geom_bar(stat = "identity", fill = "orange") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) + 
    xlab("Event Type") + ylab("Fatalities") + ggtitle("Fatalities by top 10 Weather Event Types") +
    theme(plot.title = element_text(size = 10))

p2 = ggplot(Top10_Injuries, aes(x = EVTYPE, y = INJURIES, theme_set(theme_bw()))) + 
    geom_bar(stat = "identity", fill = "pink") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) + 
    xlab("Event Type") + ylab("Injuries") + ggtitle("Injuries by top 10 Weather Event Types") +
    theme(plot.title = element_text(size = 10))

## Plot both side by side using gridExtra package
grid.arrange(p1, p2, ncol = 2, top = "Most Harmful Events with Respect to Population Health")

```

Figure 1: The weather event responsbile for the highest fatalities and injuries is the ‘Tornado’

### QUESTION 2. Across the United States, which types of events have the greatest economic consequences?

An analysis of the weather events responsible for the greatest economic consequences

Hypothesis: Economic consequences means damages. The two significant types of damage typically caused by weather events include 'properties and crops'

2.1 Data Exploration & Findings ...

Upon reviewing the column names, the property damage(PROPDMG) and crop damage(CROPDMG) columns both have another related column titled 'exponents' (i.e - PROPDMGEXP and CROPDMGEXP respectively). 

As a result, let's convert the exponent columns into numeric data for the calculation of total property and crop damages encountered.

2.2 Defining & Calcuating [ Property Damage ]

Property damage exponents for each level listed out & assigned those values for the property exponent data. 

Invalid data was excluded by assigning the value as '0'.

Then, the property damage value was calculated by multiplying the property damage and property exponent value.

```{r,prop $ crop damages analysis, echo=TRUE}

unique(strmdata$PROPDMGEXP)

#Assigning values for the property exponent strmdata 
strmdata$PROPEXP[strmdata$PROPDMGEXP == "K"] <- 1000
strmdata$PROPEXP[strmdata$PROPDMGEXP == "M"] <- 1e+06
strmdata$PROPEXP[strmdata$PROPDMGEXP == ""] <- 1
strmdata$PROPEXP[strmdata$PROPDMGEXP == "B"] <- 1e+09
strmdata$PROPEXP[strmdata$PROPDMGEXP == "m"] <- 1e+06
strmdata$PROPEXP[strmdata$PROPDMGEXP == "0"] <- 1
strmdata$PROPEXP[strmdata$PROPDMGEXP == "5"] <- 1e+05
strmdata$PROPEXP[strmdata$PROPDMGEXP == "6"] <- 1e+06
strmdata$PROPEXP[strmdata$PROPDMGEXP == "4"] <- 10000
strmdata$PROPEXP[strmdata$PROPDMGEXP == "2"] <- 100
strmdata$PROPEXP[strmdata$PROPDMGEXP == "3"] <- 1000
strmdata$PROPEXP[strmdata$PROPDMGEXP == "h"] <- 100
strmdata$PROPEXP[strmdata$PROPDMGEXP == "7"] <- 1e+07
strmdata$PROPEXP[strmdata$PROPDMGEXP == "H"] <- 100
strmdata$PROPEXP[strmdata$PROPDMGEXP == "1"] <- 10
strmdata$PROPEXP[strmdata$PROPDMGEXP == "8"] <- 1e+08

# Assigning '0' to invalid exponent strmdata
strmdata$PROPEXP[strmdata$PROPDMGEXP == "+"] <- 0
strmdata$PROPEXP[strmdata$PROPDMGEXP == "-"] <- 0
strmdata$PROPEXP[strmdata$PROPDMGEXP == "?"] <- 0

# Calculating the property damage value
strmdata$PROPDMGVAL <- strmdata$PROPDMG * strmdata$PROPEXP


# 2.3 Defining & Calcuating [ Crop Damage ]

## Crop damage exponents for each level listed out & assigned those values for the crop exponent data. 
## Invalid data was excluded by assigning the value as '0'. 
## Then, the crop damage value was calculated by multiplying the crop damage and crop exponent value.

unique(strmdata$CROPDMGEXP)

# Assigning values for the crop exponent strmdata 
strmdata$CROPEXP[strmdata$CROPDMGEXP == "M"] <- 1e+06
strmdata$CROPEXP[strmdata$CROPDMGEXP == "K"] <- 1000
strmdata$CROPEXP[strmdata$CROPDMGEXP == "m"] <- 1e+06
strmdata$CROPEXP[strmdata$CROPDMGEXP == "B"] <- 1e+09
strmdata$CROPEXP[strmdata$CROPDMGEXP == "0"] <- 1
strmdata$CROPEXP[strmdata$CROPDMGEXP == "k"] <- 1000
strmdata$CROPEXP[strmdata$CROPDMGEXP == "2"] <- 100
strmdata$CROPEXP[strmdata$CROPDMGEXP == ""] <- 1

# Assigning '0' to invalid exponent strmdata
strmdata$CROPEXP[strmdata$CROPDMGEXP == "?"] <- 0

# calculating the crop damage 
strmdata$CROPDMGVAL <- strmdata$CROPDMG * strmdata$CROPEXP


```

```{r,Crop and property damages, echo=TRUE}

# 2.4 Property Damage Summary

## Procedure = aggregate the property damage by the event type and sort the output it in descending order

prop <- aggregate(PROPDMGVAL~EVTYPE,data=strmdata,FUN=sum,na.rm=TRUE)

prop <- prop[with(prop,order(-PROPDMGVAL)),]

prop <- head(prop,10)

print(prop)


# Q2.5 Crop Damage Summary

## Procedure = aggregate the crop damage by the event type and sort the output it in descending order

crop <- aggregate(CROPDMGVAL~EVTYPE,data=strmdata,FUN=sum,na.rm=TRUE)

crop <- crop[with(crop,order(-CROPDMGVAL)),]

crop <- head(crop,10)

print(crop)

```

```{r,plot of crop & prop damages, echo=TRUE}

## Plot using ggplot2 for crop
p1 <- ggplot(crop, aes(x = EVTYPE, y = CROPDMGVAL, theme_set(theme_bw()))) +
    geom_bar(stat = "identity", fill = "green") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("Event Type") + ylab("Total Damage in $USD") + ggtitle("Total  Crop Damage")

## Plot using ggplot2 for property
p2 <- ggplot(prop, aes(x = EVTYPE, y = PROPDMGVAL, theme_set(theme_bw()))) +
    geom_bar(stat = "identity", fill = "green") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("Event Type") + ylab("Total Damage in $USD") + ggtitle("Total Property  Damage")

## Plot both side by side using gridExtra package
grid.arrange(p1, p2, ncol = 2, top = "Total Property & Crop Damage by top 10 Weather Events")

```

Figure 2: ‘Floods’ are responsbile for the highest property damage while ‘droughts’ cause the greatest crop damage.

Summary of Conclusions
Tornados are responsible for the maximum number of fatalities and injuries, followed by Excessive Heat for fatalities and Thunderstorm wind for injuries.

Floods are responsbile for maximum property damage, while Droughts cause maximum crop damage. Second major events that caused the maximum damage was Hurricanes/Typhoos for property damage and Floods for crop damage.

